---
title: "Exploratory analysis of exported cases"
author: "Sangeeta Bhatia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  word_document:
    fig_width: 7
    fig_height: 5
    fig_caption: true
    highlight: "tango"
params:
  ref_country: "sgp"
  month_col: "Total"
  exclude_repatriated: FALSE
---

Note that this analysis relies on data from IATA that are not included 
in this repository due to confidentiality restrictions.

```{r air-travel-volume-1, echo = FALSE }
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dev = c("pdf", "png"),
  dpi = 350,
  width = 15 / 2.5,
  height = 10 / 2.5,
  ## dev.args = list(pointsize = 6),
  fig.path = "figures/"
)
source("likelihood.R")
library(dplyr)
library(ggplot2)

palette <- c(
  Asia = "#E69F00",
  Europe = "#56B4E9",
  Africa = "#009E73",
  `North America` = "#0072B2",
  Oceania = "#D55E00",
  `South America` = "#CC79A7"
)

## To run this file, set the following parameters at the top of the file: ref_country and exclude_repatriated
## ref country must be one of the following: "sgp", "can", ind", "swe", "npl", "lka" or "all".
## exclude_repatriated must be TRUE or FALSE

possible_ref_countries <- list(
  sgp = "SGP",
  can = "CAN",
  ind = "IND",
  swe = "swe",
  npl = "NPL",
  lka = "LKA",
  all = c("IND", "SWE", "NPL", "LKA", "PHL", "SGP", "CAN", "FIN")
)

ref_countries <- possible_ref_countries[[params$ref_country]]

## This is from IATA data, not included in the repository
travel_vol <- readxl::read_excel("travel_volume.xlsx")

```
These are outbound flows from Wuhan for Jan to March 2016.

```{r air-travel-volume-2 }

## month_col is the column that contains months
travel_vol2 <- dplyr::select(
  travel_vol, states,
  travel_vol = params$month_col
)
## Excluding China
travel_vol2 <- travel_vol2[travel_vol2$states != "China", ]

## 18022020 We don't need the scaling factor actually, because it makes
## no difference to what we are doing.
scaling_factor <- 1
travel_vol2$travel_vol <- travel_vol2$travel_vol * scaling_factor

```


```{r air-travel-volume-3 }

travel_vol2$iso3c <- countrycode::countrycode(
  travel_vol2$states, "country.name", "iso3c"
)
```
The countries whose ISO3C code is NA here are: 

```{r air-travel-volume-4 }
dplyr::filter(travel_vol2, is.na(iso3c))
```


## Cases detected overseas



```{r air-travel-volume-5 }
## These are the exported cases -
## see https://github.com/mrc-ide/COVID19_surveillance_sensitivity/blob/master/exported_cases.csv
exported <- readr::read_csv("exported_cases.csv")
overall <- nrow(exported)
```

Here we consider cases that were not locally transmitted. 
We define local transmission as any transmission that occurred outside 
of China (Hong Kong and Macau as considered outside of China here).
We only look at cases where local transmission did not happen. 
That is, at cases with travel history to China.

```{r air-travel-volume-6 }

exported <- dplyr::filter(
  exported,
  local_transmission_y_n %in% c("n", "n - implied", "no")
)

exported$travel_history_to_hubei_y_n <- dplyr::case_when(
  exported$travel_history_to_hubei_y_n %in% c("y", "yes") ~ "y",
  exported$travel_history_to_hubei_y_n %in% c("no", "n", "n - implied") ~ "n",
  TRUE ~ exported$travel_history_to_hubei_y_n
)

repatriated <- dplyr::filter(
  exported,
  mode_of_transport_plain_train == "air - repatriated",
)
```

That is `r nrow(exported)` cases out of `r overall` total cases 
recorded at the moment. 

```{r}

travel_history <-
  dplyr::count(exported, country, travel_history_to_hubei_y_n) %>%
  tidyr::spread(key = travel_history_to_hubei_y_n, n, fill = 0)

travel_history <- dplyr::rename(
  travel_history,
  Country = country,
  `No Travel History to Hubei` = n,
  `Travel History to Hubei` = y
)
```

```{r}
knitr::kable(travel_history)
```




```{r air-travel-volume-8 }

air_codes <- c(
  "air",
  "air (inferred)",
  "air - repatriated",
  "air, train",
  "air, bus",
  "air - inferred", "air - implied"
)

exported <- exported[exported$mode_of_transport_plain_train %in% air_codes, ]

if (params$exclude_repatriated) {
  exported <- exported[!(exported$mode_of_transport_plain_train %in% c("air - repatriated")), ]
}
```
`nrow(exported)` are included in the analysis. 
The earliest date of
travel is `r min(exported$date_report_dd_mm_yyyy)`.

The latest date of travel is `r max(exported$date_report_dd_mm_yyyy)`.


```{r air-travel-volume-9 }
by_country <- dplyr::count(exported, country)
```



```{r air-travel-volume-11 }
## To get the iso3c code
by_country$country[by_country$country ==  "HK SAR"] <- "Hong Kong"

by_country$iso3c <- countrycode::countrycode(
  by_country$country, "country.name", "iso3c"
)
```


```{r}
both <- dplyr::left_join(
  x = travel_vol2,
  y = by_country,
  by = "iso3c"
)

both$monthly_average <- both$travel_vol / 3

continents <- readr::read_csv(
  "country-and-continent-codes-list-csv_csv.csv"
)

both <- dplyr::left_join(
  both, continents,
  by = c("iso3c" = "Three_Letter_Country_Code")
)
## Russian Federation is present as a European country as well as
## Asian country, removing 1.
both <- dplyr::filter(both, ! (states == "Russian Federation" & Continent_Name == "Asia"))
## Same for Tueky, Kazakhstan and Azerbaijan
both <- dplyr::filter(both, ! (states %in% c("Turkey", "Kazakhstan", "Azerbaijan")
  & Continent_Name == "Asia"))

both <- both[!is.na(both$iso3c), ]
no_cases <- which(is.na(both$n))
both$n[no_cases] <- 0

```

We assume that the number of cases detected in Singapore $X_{SGP}$ is 
distributed according to a Poisson distribution with a mean that
depends linearly on the flow from Wuhan to Singapore. 

$$X_{SGP} \sim Poisson(beta * travel_{sgp})$$

The number of cases detected in a country $I$ is similarly
distributed according to a Poisson distribution with a mean that
depends linearly on the flow from Wuhan to $I$, and a multiplicative
factor $s_e$ which is a measure of the sensitivity of the surveillance
relative to Singapore.

$$
X_{I} \sim Poisson(s_e * beta * travel_{I})
$$

## Maximum Likelihood Estimates for lambda and se


```{r air-travel-volume-14}


others <- both$iso3c[!both$iso3c %in% ref_countries]


est_with_ci <- purrr::map_dfr(
  others,
  function(other) {
    message(other)

    ll_profile <- likelihood(both, ref_countries, other)
    out <- extract_ci(ll_profile)
  }
)

est_with_ci <- dplyr::left_join(
  est_with_ci, both,
  by = "iso3c"
)
```




## Sensitivity of surveillance relative to Singapore

```{r air-travel-volume-15 }

both$slopes <- both$n / both$`travel_vol`
sgp_slope <- both$slopes[both$iso3c == "SGP"]

both$relative_slopes <- both$slopes / sgp_slope

est_with_ci$iso3c <- factor(
  est_with_ci$iso3c,
  levels = est_with_ci$iso3c[order(est_with_ci$se_mle, decreasing = TRUE)]
)

est_with_ci$n <- as.integer(est_with_ci$n)

est_with_ci <- dplyr::mutate_if(
  est_with_ci, is.numeric, round,
  digits = 6
)
```

```{r}

outfile <- "estimated_se_with_ci_ref.csv"

est_with_ci <-
  dplyr::arrange(est_with_ci, desc(se_ci_high), states)


readr::write_csv(x = est_with_ci, path = outfile)
```


## Expected number of cases overseas

```{r}

travel_ref <- both$travel_vol[both$iso3c %in% ref_countries]
travel_obs <- both$n[both$iso3c %in% ref_countries]
lhat <- sum(travel_obs) / sum(travel_ref)

lrange <- seq(from = lhat / 1000, to = 0.004, by = 0.000001)
out <- sapply(lrange, function(x) ll3(x, 0, travel_ref, 0, travel_obs, 0))

idx <- which(abs(max(out) - out) <= 3.84 / 2)

lambda_ci_low <- min(lrange[idx])
lambda_ci_high <- max(lrange[idx])

expected_case_count <- data.frame(
  lhat = lhat,
  llow = lambda_ci_low,
  lhigh = lambda_ci_high,
  traffic = both$travel_vol,
  continent = both$Continent_Name,
  country = both$states,
  obs = both$n,
  expected = lhat * both$travel_vol, 
  expected_low = lambda_ci_low * both$travel_vol,
  expected_high = lambda_ci_high * both$travel_vol,
  stringsAsFactors = FALSE
)

```

```{r}
df <- expected_case_count[order(expected_case_count$traffic, decreasing = TRUE), ]

df <- data.frame(
  Country = df$country,
  `Expected case count` = paste0(
    round(df$expected, 1),
    " (",
    round(df$expected_low, 1),
    ", ",
    round(df$expected_high, 1),
    ")"
  )
)
```

```{r}
knitr::kable(df)
```


```{r}
expected_case_count <-
  expected_case_count[order(expected_case_count$traffic, decreasing = TRUE), ]

```


```{r fig2}
expected_case_count$labels <- expected_case_count$country
expected_case_count$labels[expected_case_count$expected < 30] <- ""
ymax <- 100

ggplot(
    expected_case_count,
    aes(obs, expected, col = continent)
) + geom_point(size = 2) +
    ##geom_linerange(aes(ymin = q025, ymax = q975)) +
    theme_classic() +
    ggrepel::geom_text_repel(
        aes(x = obs, y = expected, label = labels, col = continent)
    ) +
    xlab("Observed number of cases detected outside mainland China") +
    ylab("Expected  number of cases detected outside mainland China") +
    theme(legend.position = "none",
          axis.text = element_text(size = rel(1.5))) +
    geom_abline(slope = 1, intercept = 0, alpha = 0.1) +
    scale_color_manual(values = palette, aesthetics = c("color", "fill")) +
    xlim(0, ymax) +
  ylim(0, ymax)

```
